---
title: "Reproducible Research Peer Assessment 1"
output: html_document
---



##Part 1: Load the data

Here we will load the data from a zip file (activity.zip) in the working directory into R and prints the first five lines to demonstrate they exist.

```{r}
data <- read.csv(unzip("activity.zip"))
head(data)
```

##Part 2: Total Number of Steps & Mean/Median - Wording Revist!

In this part, we will sum the total number of steps taken in on each day and create a histogram to show the frequency of steps per day 

```{r}
library(ggplot2)
totSteps <- aggregate(data$steps ~ data$date, data = data, FUN = sum)
names(totSteps) <- c("date", "steps")
ggplot(totSteps, aes(x = totSteps$steps)) + geom_histogram() + xlab("Total number of steps taken per day") +ggtitle("Total Steps Taken")
```

Now we will calculate and show the mean and median for the total number of steps taken per day
```{r}
mean(totSteps$steps)
median(totSteps$steps)
```

##Part 3: Average Daily Activity Pattern - Needs labeling

We will make a time series plot of the 5 minute interval and average number of steps taken on all days. 

```{r}
avgAct <- aggregate(data$steps ~ data$interval, data = data, FUN = mean)
names(avgAct) <- c("interval", "avgSteps")
plot(x = avgAct$interval, y = avgAct$avgSteps, type = "l")
```

Now let us determine which 5 minute interval tends to have the most activity
```{r}
avgAct[which.max(avgAct$avgSteps),]
```

